---
title: "Sovereign, getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting_started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(rmarkdown.html_vignette.check_title = FALSE)
```


As the sovereign package is under active development, its **API is not guaranteed to be stable**. As a result, this introductory vignette will remain fairly barebones for the time being, and is meant to simply and succinctly demonstrate the most up-to-date workflow supported by the sovereign package. 

## 0. Environment

```{r setup}
# load packages
suppressPackageStartupMessages(library(sovereign))     # analysis
suppressPackageStartupMessages(library(quantmod))      # FRED api
suppressPackageStartupMessages(library(dplyr))         # general cleaning
suppressPackageStartupMessages(library(lubridate))     # date functions
```

## 1. Create Data

```{r}
# pull and prepare data from FRED
quantmod::getSymbols.FRED(
    c('UNRATE','INDPRO','GS10'), 
    env = globalenv())

Data = cbind(UNRATE, INDPRO, GS10)

Data = data.frame(Data, date = zoo::index(Data)) %>%
    dplyr::filter(lubridate::year(date) >= 1990) %>% 
    na.omit()

```

## 2. Assign Regimes

```{r}

# create a regime explicitly   
Data.threshold = Data %>%
    mutate(mp = if_else(GS10 > median(GS10), 1, 0))

# assign regimes based on unsurpervised kmeans
#  (will not be used further)
regimes = 
    learn_regimes(
        data = Data, 
        regime.n = 3, 
        engine = 'kmeans')

```

## 3. Single-Regime Analaysis 

```{r}

#------------------------------------------
# single-regime var
#------------------------------------------
# estimate VAR
var =
    VAR(
        data = Data,
        p = 1,
        horizon = 10,
        freq = 'month')

# estimate IRF
irf =
    var_irf(
        var,
        bootstraps.num = 10,
        CI = c(0.05,0.95))

# estimate forecast error variance decomposition
fevd =
    var_fevd(
        var,
        horizon = 10)

#------------------------------------------
# single-regime local projection
#------------------------------------------
# estimate IRF
 lp.irfs =
    lp_irf(
      data = Data,
      shock = 'INDPRO',
      target = 'GS10',
      horizons = 20,
      lags = 2)

```

## 4. Multi-Regime Analysis

```{r}

#-------------------------------------------
# multi-regime var
#------------------------------------------
# estimate multi-regime VAR
tvar =
    threshold_VAR(
        data = Data.threshold,
        regime = 'mp',
        p = 1,
        horizon = 1,
        freq = 'month')

# estimate IRF
tvar.irf =
    threshold_var_irf(
        tvar,
        horizon = 10,
        bootstraps.num = 10,
        CI = c(0.05,0.95))

# estimate forecast error variance decomposition
tvar.fevd =
    threshold_var_fevd(
        tvar,
        horizon = 10)

#-------------------------------------------
# multi-regime local projection
#------------------------------------------
# estimate IRF
tlp.irfs =
  threshold_lp_irf(
    data = dplyr::select(Data.threshold, -mp),
    thresholdVar = dplyr::select(Data.threshold, mp, date),
    shock = 'INDPRO',
    target = 'GS10',
    horizons = 20,
    lags = 2)


```
